---
title: "Introduction to mdeinR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to mdeinR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
```

## What is mdeinR?

`mdeinR` measures **Memorable Dining Experiences (MDE)** from free-text
restaurant reviews. It scores each review sentence across five dimensions:

| Dimension | What it captures |
|---|---|
| **sensory** | Aroma, ambiance, visual presentation, texture |
| **affect** | Emotions — joy, amazement, enchantment |
| **social** | Warmth, caring, togetherness, celebration |
| **intellectual** | Curiosity, learning, discovery, knowledge |
| **behavioral** | Engagement, activities, interactive experiences |

The scoring algorithm is inspired by
[sentimentr](https://github.com/trinker/sentimentr): each matched keyword
scores `1 / sentence_word_count`, then a context window adjusts for
valence shifters (negators, amplifiers, de-amplifiers, adversatives).

---

## Installation

```r
devtools::install_github("gvajpai/mdeinR")
```

After installing, build the data objects once from the package source:

```r
source("data-raw/build_dictionary.R")
```

---

## The sample dataset

`mdeinR` ships with `restaurant_reviews` — 400 reviews across eight
fictional restaurants, designed to demonstrate every package feature.

```{r load-data}
library(mdeinR)

# Inspect the sample dataset
head(restaurant_reviews, 4)
```

```{r data-summary}
# Star rating distribution
table(restaurant_reviews$stars)

# Reviews per restaurant
table(restaurant_reviews$restaurant)
```

---

## Step 1 — Basic MDE scoring with `mde()`

Pass a character vector to `mde()` to get sentence-level scores.

```{r basic-mde}
# Score a handful of reviews
sample_reviews <- c(
  "The aroma and presentation were absolutely stunning.",
  "Staff were warm and genuinely caring — we felt completely at home.",
  "We discovered exotic spices and learned about Moroccan culinary history.",
  "The tableside performance by the chef was mesmerising and engaging.",
  "We celebrated our anniversary here and felt truly joyful and special."
)

mde(sample_reviews)
```

Each row is one sentence. The scores are `1 / word_count` per matched
dimension keyword, adjusted by any valence shifters in the context window.

---

## Step 2 — Valence shifters in action

### Negation

A negator (`not`, `never`, `didn't`, etc.) within 5 words before or 2 words
after a dimension keyword flips its score negative — which is then clamped
to zero, since MDE theory only counts positive occurrences.

```{r negation}
# Positive vs. negated versions of the same sentence
mde(c(
  "The aroma was stunning and the staff were incredibly warm.",
  "The aroma was not stunning and the staff were not warm."
))
```

Notice how the negated review scores zero — the `not` cancels the
dimension words.

### Amplification

Amplifiers (`absolutely`, `incredibly`, `extremely`, etc.) multiply the
raw score by `1 + amplifier.weight` (default 0.8), so an amplified keyword
scores up to 1.8x more than an unamplified one.

```{r amplification}
mde(c(
  "The ambiance was beautiful.",           # base score
  "The ambiance was absolutely beautiful." # amplified
))
```

### De-amplification

De-amplifiers (`somewhat`, `slightly`, `fairly`, etc.) reduce the score by
`1 - n.neutral` (default 0.2), giving a 20% reduction.

```{r deamplification}
mde(c(
  "The ambiance was beautiful.",          # base
  "The ambiance was somewhat beautiful."  # de-amplified
))
```

### Adversative conjunctions

Words like `but` and `however` signal a contrast. When an adversative
appears **before** a dimension keyword in the context window, the keyword
receives a reduced score (`1 - n.neutral`), reflecting lower certainty for
the positive claim following the contrast.

```{r adversative}
mde(c(
  "The aroma was stunning and the staff were warm.",   # no adversative
  "The aroma was stunning but the staff were warm."    # adversative before 'warm'
))
```

The `social` score for `warm` is slightly reduced in the second sentence
because `but` appeared before it.

---

## Step 3 — Scoring the full sample dataset

```{r score-all}
# Score all 400 reviews — returns one row per sentence
scores <- mde(restaurant_reviews$text)
dim(scores)
head(scores)
```

```{r score-summary}
# Average scores across all reviews
colMeans(scores[, c("sensory", "affect", "behavioral", "social", "intellectual")])
```

---

## Step 4 — Aggregate by group with `mde_by()`

`mde_by()` averages sentence-level scores within groups. Pass a
`data.frame` with a `by` column.

```{r mde-by-restaurant}
# Average MDE scores by restaurant
by_restaurant <- mde_by(restaurant_reviews, by = "restaurant")
by_restaurant[, .(restaurant, sensory, affect, social, intellectual, behavioral, n_sentences)]
```

```{r mde-by-stars}
# Average MDE scores by star rating
by_stars <- mde_by(restaurant_reviews, by = "stars")
by_stars[, .(stars, sensory, affect, social, intellectual, behavioral, n_sentences)]
```

You can also group by multiple columns simultaneously:

```{r mde-by-multi}
# Group by both restaurant and star rating
by_both <- mde_by(restaurant_reviews, by = c("restaurant", "stars"))
head(by_both[, .(restaurant, stars, sensory, affect, social, n_sentences)])
```

---

## Step 5 — `get_sentences()` for multi-sentence reviews

For long reviews with multiple sentences, `get_sentences()` splits them
first so each sentence is scored independently.

```{r get-sentences}
# A multi-sentence review
long_review <- "The aroma was absolutely stunning from the moment we entered.
                Staff were incredibly warm and caring throughout the evening.
                We learned fascinating techniques from the chef's tableside demonstration.
                The food was not particularly amazing but the overall experience was joyful."

# Split into sentences
sents <- get_sentences(long_review)
print(sents)
```

```{r score-sentences}
# Score each sentence
mde(sents)
```

Notice that the fourth sentence scores lower — the negator `not` cancels
`amazing`, and the adversative `but` reduces the `affect` score for `joyful`.

---

## Step 6 — Highlighting dimension keywords

`highlight_mde()` annotates a review with XML-style dimension tags for
qualitative inspection.

```{r highlight}
highlight_mde("The aroma was stunning and staff were so warm and caring.")
```

---

## Step 7 — Customising valence shifters

Add domain-specific negators or amplifiers with `update_valence_shifters()`.

```{r custom-vs}
# Add restaurant-specific terms as negators
custom_vs <- update_valence_shifters(
  data.frame(
    x = c("overpriced", "undercooked", "bland", "disappointing"),
    y = c(1L, 1L, 1L, 1L)   # type 1 = negator
  )
)

# Compare with and without custom shifters
mde("The food was amazing but overpriced.",
    valence_shifters_dt = custom_vs)

mde("The food was amazing but overpriced.")
```

---

## Step 8 — Inspecting the MDE dictionary

```{r dictionary}
# Number of words per dimension
table(mde_dictionary$dimension)

# Sample words from each dimension
head(mde_dictionary[mde_dictionary$dimension == "sensory", ])
head(mde_dictionary[mde_dictionary$dimension == "intellectual", ])
```

---

## Scoring algorithm summary

For each sentence:

1. Tokenise into lower-case words using Unicode-aware regex (punctuation
   stripped correctly — `"amazing,"` matches `"amazing"`).
2. For each token matching `mde_dictionary`, assign raw score `1 / n_words`.
3. Check a context window of 5 words before and 2 words after:
   - **Negator**: multiply by `-1` → clamped to 0
   - **Amplifier**: multiply by `1 + 0.8 = 1.8`
   - **De-amplifier**: multiply by `1 - 0.2 = 0.8`
   - **Adversative before keyword**: multiply by `0.8`
4. Sum adjusted scores per dimension across the sentence.
5. Negative scores set to 0 (MDE counts only positive occurrences).

`mde_by()` averages sentence-level scores within each group.

---

## Citation

```bibtex
@article{vajpai2025mde,
  title   = {Designing a memorable dining experience lexicon based on
             theory and text mining},
  author  = {Vajpai, Gopi Nath and Webb, Timothy and Beldona, Srikanth},
  journal = {International Journal of Hospitality Management},
  volume  = {130},
  pages   = {104245},
  year    = {2025},
  doi     = {10.1016/j.ijhm.2025.104245}
}
```
