---
title: "Introduction to mdeinR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to mdeinR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

`mdeinR` scores the five dimensions of a **Memorable Dining Experience
(MDE)** — *sensory*, *affect*, *behavioral*, *social*, and *intellectual* —
from free-text restaurant reviews.

The scoring algorithm is inspired by
[sentimentr](https://github.com/trinker/sentimentr) by Tyler W. Rinker:
reviews are split into sentences with the built-in `get_sentences()`;
every MDE keyword is scored as `1 / sentence_word_count`;
a context window around each keyword is scanned for valence shifters
(negators, amplifiers, de-amplifiers, adversative conjunctions) from the
bundled `valence_shifters` table.

The lexicon (324 words) was developed and validated by Vajpai, Webb &
Beldona (2025).

---

## Installation

```r
# Install from GitHub
devtools::install_github("gvajpai/mdeinR")
```

After cloning or installing, build the data objects once:

```r
source("data-raw/build_dictionary.R")
```

---

## Quick start

```{r, eval=FALSE}
library(mdeinR)

reviews <- c(
  "The aroma and presentation were absolutely stunning, a feast for the senses.",
  "Staff were warm and genuinely caring — we truly felt at home.",
  "We discovered exotic spices and learned about the history of Moroccan cuisine.",
  "Nothing particularly noteworthy about the food or the décor."
)
```

### Sentence-level scores

```{r, eval=FALSE}
mde(reviews)
```

### Document-level aggregate

```{r, eval=FALSE}
mde_by(reviews)
```

### Group-level aggregate

```{r, eval=FALSE}
library(data.table)

restaurant_reviews <- data.table(
  restaurant = c("A", "A", "B", "B", "B"),
  text = c(
    "The aroma and presentation were stunning.",
    "Staff were caring and we felt at home.",
    "A truly intellectual adventure through Thai cuisine.",
    "Nothing special — plain and forgettable.",
    "The ambiance was gorgeous and the music was delightful."
  )
)

mde_by(restaurant_reviews, by = "restaurant")
```

---

## Using get_sentences() first (recommended for long reviews)

For documents with multiple sentences, pass the `get_sentences()` output to
`mde()` so that sentence splitting is handled correctly:

```{r, eval=FALSE}
long_review <- "The food was amazing. Staff were incredibly caring.
                I discovered dishes I had never heard of before.
                But the service was not fast enough."

mde(get_sentences(long_review))
```

---

## Highlighting MDE keywords

`highlight_mde()` wraps matched tokens in XML-style tags for qualitative
inspection:

```{r, eval=FALSE}
highlight_mde("The aroma was wonderful and staff were so caring.")
# [1] "<sensory>aroma</sensory> was wonderful and staff were so <social>caring</social>."
```

---

## Valence shifters

Negators (`not`, `never`, `don't`, ...), amplifiers (`very`, `extremely`, ...),
de-amplifiers (`somewhat`, `slightly`, ...), and adversative conjunctions
(`but`, `however`, ...) are handled automatically via the bundled
`mdeinR::valence_shifters` table.

You can extend this table with domain-specific terms:

```{r, eval=FALSE}
custom_vs <- update_valence_shifters(
  data.frame(
    x = c("overpriced", "undercooked", "bland"),
    y = c(1L, 1L, 1L)   # treat as negators
  )
)

mde("The food was amazing but overpriced.", valence_shifters_dt = custom_vs)
```

---

## Inspecting the dictionary

```{r, eval=FALSE}
# All 324 tokens
head(mde_dictionary, 10)

# Words per dimension
table(mde_dictionary$dimension)
```

---

## Citation

If you use `mdeinR` in your research, please cite:

> Vajpai, G. N., Webb, T., & Beldona, S. (2025). Designing a memorable
> dining experience lexicon based on theory and text mining.
> *International Journal of Hospitality Management*, 130, 104245.
> https://doi.org/10.1016/j.ijhm.2025.104245
